---
title: "Project 2: King County Housing Data"
author: "Cole Stetson"
date: "05/05/2026"
output: 
  bookdown::html_document2: 
    highlight: textmate
    theme: yeti
---

```{r, label = "SETUP", echo = FALSE, results= 'hide', message = FALSE, warning = FALSE}
set.seed(123)
library(knitr)
library(ggplot2)
knitr::opts_chunk$set(comment = NA,  fig.align = 'center', fig.height = 5, fig.width = 5, warning = FALSE, message = FALSE, tidy.opts=list(blank = TRUE, width.cutoff= 75))
```

# Variable Descriptions

Variable descriptions were obtained from King County, Department of Assessments. All feature engineering should be done in the first code chunks of your document.

- id - Unique ID for each home sold
- date - Date of the home sale
- price - Price of each home sold
- bedrooms - Number of bedrooms
- bathrooms - Number of bathrooms, where .5 accounts for a room with a toilet but no shower
- sqft_living - Square footage of the apartments interior living space
- sqft_lot - Square footage of the land space
- floors - Number of floors
- waterfront - A dummy variable for whether the apartment was overlooking the waterfront or not * 1’s represent a waterfront property, 0’s represent a non-waterfront property
- view - An index from 0 to 4 of how good the view of the property was, 0 - lowest, 4 - highest
- condition - An index from 1 to 5 on the condition of the apartment, 1 - lowest, 4 - highest
- grade - An index from 1 to 13, where 1-3 falls short of building construction and design, 7 has an average level of construction and design, and 11-13 have a high quality level of construction and design.
- sqft_above - The square footage of the interior housing space that is above ground level
- sqft_basement - The square footage of the interior housing space that is below ground level
- yr_built - The year the house was initially built
- yr_renovated - The year of the house’s last renovation
- zipcode - What zipcode area the house is in
- lat - Lattitude
- long - Longitude
- sqft_living15 - The square footage of interior housing living space for the nearest 15 neighbors
- sqft_lot15 - The square footage of the land lots of the nearest 15 neighbors

# Data and Feature Engineering

## Data Set (housedata) and Data to send predictions (housedataT)

```{r}
# Data Set

housedata <- read.csv("https://lasanthi-asu.github.io/STT3851ClassRepo/Rmarkdown/Data/housedata.csv", 
                      colClasses = c(id = "character", date = "character", 
                                     yr_built = "character", zipcode = "factor", grade = "factor"))

housedata$date <- as.Date(housedata$date, "%Y%m%d")
housedata$waterfront <- factor(housedata$waterfront, labels = c("No", "Yes"))
housedata$condition <- factor(housedata$condition, labels = c("poor", "fair", "average", "good", "very good"))
housedata$yr_renovated <- ifelse(housedata$yr_renovated == 0, housedata$yr_built, housedata$yr_renovated)
housedata$yr_built <- as.Date(ISOdate(housedata$yr_built, 9, 1))  # Complete Year, Sept 1
housedata$yr_renovated <- as.Date(ISOdate(housedata$yr_renovated, 9, 1))  # Last renovated Year, Sept 1
housedata <- housedata[, -1]


# Test Data Set (only use this to send predictions)

#### Perform same steps with test set

housedataT <- read.csv("https://lasanthi-asu.github.io/STT3851ClassRepo/Rmarkdown/Data/housedataTEST.csv", 
                      colClasses = c(id = "character", date = "character", 
                                     yr_built = "character", zipcode = "factor", grade = "factor"))

housedataT$date <- as.Date(housedataT$date, "%Y%m%d")
housedataT$waterfront <- factor(housedataT$waterfront, labels = c("No", "Yes"))
housedataT$condition <- factor(housedataT$condition, labels = c("poor", "fair", "average", "good", "very good"))
housedataT$yr_renovated <- ifelse(housedataT$yr_renovated == 0, housedataT$yr_built, housedataT$yr_renovated)
housedataT$yr_built <- as.Date(ISOdate(housedataT$yr_built, 9, 1))  # Complete Year, Sept 1
housedataT$yr_renovated <- as.Date(ISOdate(housedataT$yr_renovated, 9, 1))  # Last renovated Year, Sept 1
housedataT <- housedataT[, -1]
```

```{r}
library(DT)
datatable(housedata[, ], rownames = FALSE)
```

# Where to Start

Correlation Matrix maybe helpful.

Here is an example of a correlation matrix with some of the predictors


```{r}
# Corrplot
library(corrplot)
housedata1 <- housedata[, c(2:7, 9, 12, 13, 17:20)]
cor(housedata1)
```

```{r}
# Corrplot
library(corrplot)
corrplot(cor(housedata1))
```

**Analysis of Correlation Matrix/Plot:** 

 - Strongest Correlation with Price: sqft_living, sqft_above, sqft_living15, bathrooms, and view.
 
 - Strongest Interaction: All of the sqft variables with each other and with bathrooms/bedrooms.

# Building Models

## Most basic model (named mod.zip below).

```{r}
mod.zip <- lm(price ~ 1, data = housedata)
summary(mod.zip)
```

**EXAMPLE/TEMPLATE**
 
```{r}
set.seed(1)
mymod1 <- lm(price ~ ., data= housedata)
netnon_dole <- predict(mymod1, newdata= housedataT)
plot(mymod1)
```

**Analysis:** Expected skewness in residuals - needs log transformation on price. Curve in scale-location - needs nonlinear pieces.

**Organizing Data and Variables - Test with lm**

```{r}
str(housedata)
```

**Analysis:** There's a mix of numerical variables, categorical variables, and variables that are labeled "Date". The four qualitative variables are waterfront, condition, grade, and zipcode. The three date variables are date, yr_built, and yr_renovated. I want to use a Ridge model as my final prediction model because it's good for highly correlated variables and I see a lot of correlation in the matrix/plot. Because of this I don't want to actually remove any variables because I know Ridge can use all of them and minimize them accordingly. The date predictors seem of kind of useless right now because a date alone doesn't really mean anything statistically. Therefore I want to clean up the "Date" predictors so that they are more useful but still included. I want to add a new numeric "age" variable to replace yr_built and date so that the difference (the actual current age of the house) is used. I also want to do the same thing with yr_renovated and add a numeric "age_renovated" variable so that the measure of how currently the house has been renovated is used. I think that both of these engineered features will make a noticeable difference. To futher improve the model, I will test out a bunch of different interaction terms by adding them and then choosing to either keep or remove them after seeing how the MSE changes. To see MSE I am going to split the datatset "housedata" into a training set and test set for cross validation. 

```{r}
library(glmnet)
set.seed(123)

#Adding new predictors to both housedata and housedataT
housedata$age <- 2026 - as.numeric(format(housedata$yr_built, "%Y"))
housedata$age_renovated <- 2026 - as.numeric(format(housedata$yr_renovated, "%Y"))
#Note: I know that we did not learn the "as.numeric(format())" in our class. I looked up online how to extract the year from the date and this is was what I found. This is not completely original work but it was an original idea and I thought using external resources would be okay for this situation.
housedata$log_sqft_living <- log(housedata$sqft_living) #difference between 1000-2000 sqft has bigger impact on price than 5000-6000 sqft
housedata$log_sqft_lot <- log(housedata$sqft_lot) #same concept as last line
housedata$bathrooms_sq <- housedata$bathrooms^2 
housedata$living_grade <- housedata$sqft_living * as.numeric(housedata$grade)
#I am using the same "as.numeric() function on the categorical variable grade to make it numerical for interaction. I think house size and quality of construction is an important aspect to capture because some big houses are built poorly. 
housedata$living_view <- housedata$sqft_living * housedata$view #big homes that also have a good view have a bigger impact on price
housedata$bed_bath <- housedata$bedrooms * housedata$bathrooms
housedata$age_condition <- housedata$age * as.numeric(housedata$condition)
#using as.numeric() function on condition for same thing. I think that some old houses have been properly maintained or renovated so they shouldn't be as punished for being old. 
housedata$rel_living <- housedata$sqft_living / housedata$sqft_living15 #how big a house is in relation to houses around it is also very important. My parents found a good deal on their house in Asheville because its the smallest house in a nice neighborhood. 

#Note: I tried a lot of other interaction variables during the process and removed ones I didn't find useful. These are just the ones I want to include for the final model. Im sure that there are some important ones that I am missing. The sqft interactions terms with one another just felt repetitive so I didn't try many out. I wanted to focus on thinking outside the box and find some hidden relationships that could make a big difference. 

#Additonal Post-Note: Changing the lambda made a huge difference. Honestly I could've just left every additonal variable I created because the Ridge model with a good lambda would've minmized the less usefull variables anyway.

housedataT$age <- 2026 - as.numeric(format(housedataT$yr_built, "%Y"))
housedataT$age_renovated <- 2026 - as.numeric(format(housedataT$yr_renovated, "%Y"))
housedataT$log_sqft_living <- log(housedataT$sqft_living)
housedataT$log_sqft_lot <- log(housedataT$sqft_lot)
housedataT$bathrooms_sq <- housedataT$bathrooms^2
housedataT$living_grade <- housedataT$sqft_living * as.numeric(housedataT$grade)
housedataT$living_view <- housedataT$sqft_living * housedataT$view
housedataT$bed_bath <- housedataT$bedrooms * housedataT$bathrooms
housedataT$age_condition <- housedataT$age * as.numeric(housedataT$condition)
housedataT$rel_living <- housedataT$sqft_living / housedataT$sqft_living15



#Training Splits for Cross Validation
train <- sample(1:nrow(housedata), nrow(housedata) / 2)
train_data <- housedata[train, ]
test_data  <- housedata[-train, ]


x_train <- model.matrix(price ~ . - date - yr_built - yr_renovated, data = train_data)[, -1]
x_test  <- model.matrix(price ~ . - date - yr_built - yr_renovated, data = test_data)[, -1]

y_train <- log(train_data$price)
y_test  <- log(test_data$price)

grid <- 10^seq(4, -4, length = 200)
#This felt like the biggest piece of my entire model. I had started with the 10^seq(10, -2, length=100) template that we had used in class but I noticed that Lasso and Ridge both picked lambda=0.01. I thought this was a little strange so I wanted to change how I defined the grid and see if it made a difference. I had to look up online what the roles of each number in the function were then once I understood the format I started to change them. This was probably around the 5th setup that I tried and it drastically lowered my MSE so I just stuck with it.
```

**Explanation:** I'm sorry if everything looks a little sloppy. I was trying things quickly 

```{r}
lm_model <- lm(log(price) ~ . - date - yr_built - yr_renovated, data = train_data)

lm_pred <- predict(lm_model, newdata = test_data)

lm_mse <- mean((y_test - lm_pred)^2)

lm_mse
plot(lm_model)
```

**LASSO**

```{r}
lasso_fit <- glmnet(x_train, y_train, alpha = 1, lambda=grid)
cv_lasso <- cv.glmnet(x_train, y_train, alpha = 1, lambda= grid)

best_lambda_lasso <- cv_lasso$lambda.min
best_lambda_lasso

lasso_model <- glmnet(x_train, y_train, alpha = 1, lambda = best_lambda_lasso)

coef(lasso_model)
```

```{r}
lasso_pred <- predict(cv_lasso, s = best_lambda_lasso, newx = x_test)

lasso_mse <- mean((y_test - lasso_pred)^2)
lasso_mse

plot(lasso_pred, y_test - lasso_pred)
```

Analysis: Higher MSE than OLS which shows that removing some variables entirely may not be the best idea.

**RIDGE**

```{r}
library(glmnet)

ridge_fit <- glmnet(x_train, y_train, alpha = 0, lambda=grid)
cv_ridge <- cv.glmnet(x_train, y_train, alpha = 0, lambda= grid)

best_lambda_ridge <- cv_ridge$lambda.min
best_lambda_ridge

ridge_model <- glmnet(x_train, y_train, alpha = 0, lambda = best_lambda_ridge)
```

```{r}
ridge_pred <- predict(cv_ridge, s = best_lambda_ridge, newx = x_test)

ridge_mse <- mean((y_test - ridge_pred)^2)
ridge_mse

plot(ridge_pred, y_test - ridge_pred)
```

Analysis: Lower MSE than Lasso, most likely works better because of variables being highly correlated. This confirms ideas from Scale-Location plot from OLS.

**MSE Table**

```{r}
model_results <- data.frame(Model = c("OLS", "LASSO", "Ridge"), Test_MSE = c(lm_mse, lasso_mse, ridge_mse))

model_results
```

**Format for Final Prediction "Stetson_Cole"**

```{r}

x_all <- model.matrix(price ~ . - date - yr_built - yr_renovated, data = housedata)[, -1]
y_all <- log(housedata$price)

cv_ridge_full <- cv.glmnet(x_all, y_all, alpha = 0, lambda = grid)
best_lambda_full <- cv_ridge_full$lambda.min
housedataT_final <- model.matrix(~ . - date - yr_built - yr_renovated, data = housedataT)[, -1]

Stetson_Cole <- predict(cv_ridge_full, s = best_lambda_full, newx = housedataT_final)

Stetson_Cole <- exp(Stetson_Cole)

nrow(housedataT)
length(Stetson_Cole)

write.csv(Stetson_Cole, file = "Stetson_Cole.csv", row.names = FALSE)
```


